Exploring SSA Optimization Techniques

### 2014-11-10 doc/paper.tex
Added constant branch optimization

\documentclass[12pt, a4paper]{article}
\usepackage{amsmath, listingsutf8, mathtools}
\title{Exploring SSA Optimization Techniques}
\author{A. D. Blom}
\date{}
\begin{document}
  \maketitle
  \begin{abstract}
In this era of global digitalization, the need for performant software is larger
than ever. For the production of fast software not only 
fast hardware and solidly written code are needed, but also a well-optimizing 
compiler. This paper explores some common optimization techniques based on an 
SSA intermediate language.
  \end{abstract}

@@

Once the value of b is determined, it should then also be filled in, to fold 
further. Since the value of b is 100, it can be used to fill in the return 
expression:

@@

\subsection{Considerations}
\subsubsection{Platform incompatibilities}
There is a way compiler-based constant folding might stand in the way of the 
programmer. Mostly the compiler can do this when folding away instructions 
operating on floating point operands, because different targets may compute 
floating point operations differently. Therefore cross-compilation becomes an 
issue; if a floating point instruction for target Y normally yielding Vy, it 
yields Vx when folded away by target X, causing different semantics before and 
after optimisation.

@@

>>>>

\documentclass[12pt, a4paper]{article}
\usepackage{amsmath, listingsutf8, mathtools}
\lstset{numbers=left}
\title{Exploring SSA Optimization Techniques}
\author{A. D. Blom}
\date{}
\begin{document}
  \maketitle
  \begin{abstract}
In this era of global digitalization, the need for performant software is larger
than ever. For the production of fast software not only 
fast hardware and solidly written code are needed, but also a well-optimizing 
compiler. This paper explores some common optimization techniques based on an 
SSA intermediate language.
  \end{abstract}

@@

Once the value of \verb+b+ is determined, it should then also be filled in, to fold 
further. Since the value of \verb+b+ is 100, it can be used to fill in the return 
expression:

@@

\subsection{Considerations}
\subsubsection{Platform incompatibilities}
There is a way compiler-based constant folding might stand in the way of the 
programmer. Mostly the compiler can do this when folding away instructions 
operating on floating point operands, because different targets may compute 
floating point operations differently. Therefore cross-compilation becomes an 
issue; if a floating point instruction for target $Y$ normally yielding $V_y$, it 
yields $V_x$ when folded away by target $X$, causing different semantics before and 
after optimisation.

@@

\section{Constant split removal}
\subsection{Problem}
After constant folding, some \verb+split+ instructions may branch on a constant
condition:

\begin{lstlisting}
La:	...
	split((_Bool)1, Lb, Lc);
Lb:	ret((int)0);
Lc:	ret((int)1);
\end{lstlisting}

Could be converted easily into:

\begin{lstlisting}
La:	...
	jmp(Lb);
Lb:	ret((int)0);
Lc:	ret((int)1);
\end{lstlisting}

This has only minor implications for further flow, except that it removes a
predecessor from block \verb+Lc+. The only way that that affects SSA validity is
that a block-expression pair may need to be removed from $\phi$ nodes in
\verb+Lc+.

Removing this predecessor may also have implications for further block inlining;
if a block has only one predecessor and the predecessor has only one successor,
the block could be merged with its predecessor.

\subsection{Implementation}
The implementation of this optimization simply needs to check whether the first
parameter of a \verb+split+ instruction is constant, and convert it into a
\verb+jmp+ accordingly. It also needs to check for the presence of $\phi$ nodes
in the block not covered by the \verb+jmp+ instruction, and remove them accordingly:

\begin{lstlisting}
La:	...
	split((_Bool)1, Lb, Lc);
Lb:	...
Lc:	int tA = phi(La, (int)10, ...);
	...
\end{lstlisting}

Needs to get rid of the \verb+La+ items from the \verb+tA+ $\phi$ node too:

\begin{lstlisting}
La:	...
	jmp(Lb);
Lb:	...
Lc:	int tA = phi(...);
	...
\end{lstlisting}
